
\section{Related Work}
\label{sec:related}

\textbf{IMU-based HAR.}
Deep learning for HAR has progressed from single-dataset CNN-LSTM models~\cite{deepconvlstm} to approaches addressing practical heterogeneity: Cosmo~\cite{cosmo} demonstrates contrastive fusion for small-data multimodal HAR, UniHAR~\cite{unihar} tackles cross-user and cross-device generalization via physics-informed augmentation, and ClusterFL~\cite{clusterfl} captures inter-user relationships through federated learning. Despite these advances, all remain architecturally bound to a fixed channel count, sampling rate, and closed-world label set, limiting transferability across sensing setups.

\textbf{Self-supervised pretraining and foundation models for IMU sensing.}
Self-supervised learning reduces dependence on labeled IMU data: Contrastive methods learn augmentation-invariant representations~\cite{simclr}, while masked autoencoding has been adapted from vision~\cite{mae} and language~\cite{bert} to IMU sequences. For example, LiMU-BERT~\cite{limubert} pretrains via span-masked reconstruction at 20\,Hz, and CrossHAR ~\cite{crosshar} combines masked reconstruction with contrastive learning for cross-dataset HAR. PatchTST~\cite{patchtst} demonstrates that patch-based channel-independent tokenization improves time-series modeling, a design principle we adopt. However, fixed-length positional encodings constrain variable sampling rates, and all methods require a dataset-specific classifier. Foundation models for general time-series data aim for broader coverage: MOMENT~\cite{moment} is a T5-style encoder pretrained on 13 time-series domains via masked patch reconstruction (341.2M parameters), and UniMTS~\cite{unimts} targets unified pretraining for motion time series across heterogeneous sensor configurations. In the tabular domain, Mitra~\cite{mitra} demonstrates that curated mixtures of synthetic priors can train foundation models that generalize across diverse real-world datasets via in-context learning. These general-purpose approaches trade domain-specific inductive biases for breadth; HALO occupies an intermediate position as a domain-specific foundation model for wearable sensing with HAR-specific designs and a language-based label interface.

\textbf{Language-grounded perception.}
CLIP~\cite{clip} demonstrated that aligning vision and text in a shared embedding space enables open-vocabulary zero-shot classification without per-class classifiers. ImageBind~\cite{imagebind} extends this to multiple modalities including IMU, but is not designed for cross-dataset HAR or variable sensor configurations. LanHAR~\cite{lanhar} applies CLIP-style alignment to HAR by training a sensor Transformer against LLM-generated activity descriptions. LLaSA~\cite{llasa} takes an alternative approach, leveraging LLMs directly for natural language reasoning about IMU data. Unlike LanHAR, which lacks self-supervised pretraining and assumes fixed sensor configurations, HALO integrates heterogeneity-aware pretraining with synonym-aware contrastive alignment, enabling a capability none of these prior systems achieve: a single deployable model that handles variable sampling rates, channel counts, and sensor placements for open-set HAR.
